Efficient estimation for large-scale linkage disequilibrium patterns of the human genome

Version 1

Abstract

Abstract

In this study, we proposed an efficient algorithm (X-LD) for estimating LD patterns for a genomic grid, which can be of inter-chromosomal scale or of a pair of small segments. Compared with conventional methods, the proposed method was significantly faster, and consequently we were permitted to explore in depth unknown or reveal long-anticipated LD features of the human genome. Having applied the algorithm as demonstrated in 1000 Genome Project (1KG), we found: I) The extended LD, driven by population structure, was universally existed, and the strength of inter-chromosomal LD was about 10% their respective intra-chromosomal LD in relatively homogeneous cohorts, such as FIN and to nearly 56% in admixed cohort, such as ASW. II) After splitting each chromosome into upmost more than a half million grids, we elucidated the LD of HLA region was nearly 42 folders higher than chromosome 6 in CEU and 11.58 in ASW; on chromosome 11, we observed that the LD of its centromere was nearly 94.05 folders higher than chromosome 11 in YRI and 42.73 in ASW. III) We uncovered the long-anticipated inversely proportional linear relationship between the length of a chromosome and the strength of chromosomal LD, and their Pearson’s correlation was on average over 0.80 for 26 1KG cohorts. However, this linear norm was so far perturbed by chromosome 11 given its more completely sequenced centromere region. Uniquely chromosome 8 of ASW was found most deviated from the linear norm than any other autosomes. The proposed algorithm has been realized in C++ (called X-LD) and available at https://github.com/gc5k/gear2, and can be applied to explore LD features in any sequenced populations.

Introduction

Linkage disequilibrium (LD) is the association for a pair of loci and the metric of LD serves as the basis for developing genetic applications in agriculture, evolutionary biology, and biomedical researches (Weir, 2008; Hill and Robertson, 1966). The structure of LD of the human genome is shaped by many factors, mutation, recombination, population demography, epistatic fitness, and completeness of genomic data itself (Myers et al., 2005; Nei and Li, 1973; Ardlie et al., 2002). Due to its overwhelming cost, LD structure investigation is often compromised to a small genomic region (Chang et al., 2015; Theodoris et al., 2021), and their typical LD structure is as illustrated for a small segment (Barrett et al., 2005). Now, given the availability of large-scale genomic data, such as millions of single nucleotide polymorphisms (SNPs), the large-scale LD patterns of the human genome play crucial roles in determining genomics studies, and many theories and useful algorithms upon large-scale LD structure, from genome-wide association studies, polygenic risk prediction for complex diseases, and choice for reference panels for genotype imputation (Vilhjálmsson et al., 2015; Yang and Zhou, 2020; Bulik-Sullivan et al., 2015; Yang et al., 2011; Das et al., 2016).

However, there are impediments, largely due to intensified computational cost, in both investigating large-scale LD and providing high-resolution illustration for their details. If we consider a genomic grid that is consisted of m2 SNP pairs, given a sample of n individuals and m SNPs (n ≪ m) – typically as observed in 1000 Genomes Project (1KG) (Lowy-Gallego et al., 2019), its benchmark computational time cost for estimating all pairwise LD is 𝒪(nm2), a burden that quickly drains computational resources given the volume of the genomic data. In practice, it is of interest to know the mean LD of the SNP pairs for a genomic grid, which covers mi × mj SNP pairs. Upon how a genomic grid is defined, a genomic grid consequently can be consisted of : i) the whole genome-wide m2 SNP pairs, and we denote their mean LD as ℓg; ii) the intra-chromosomal mean LD for the ith chromosome of SNP pairs, and denote as ℓi; iii) the inter-chromosomal mean LD ith and jth chromosomal mimj SNP pairs, and denoted as ℓij.

In this study we propose an efficient algorithm that can estimate ℓg, ℓi, and ℓij, the computational time of which can be reduced from to 𝒪(n2mi) for ℓi and 𝒪(nmimj) to 𝒪(n2mi + n2mj) for ℓij. The rationale of the proposed method relies on the connection between the genetic relationship matrix (GRM) and LD (Chen, 2014; Goddard, 2009), and in this study a more general transformation from GRM to LD can be established via Isserlis’s theorem (Isserlis, 1918; Zhou, 2017). The statistical properties, such as sampling variance, of the estimated LD have been derived too.

The proposed method can be analogously considered a more powerful realization for Haploview (Barrett et al., 2005), but additional utility can be derived to bring out unprecedented survey of LD patterns of the human genome. As demonstrated in 1KG, we consequently investigate how biological factors such as population structure, admixture, or variable local recombination rates can shape large-scale LD patterns of the human genomes.

The proposed method provides statistically unbiased estimates for large-scale LD patterns and shows computational merits compared with the conventional methods (Figure 2).

We estimated ℓg, and 22 autosomal ℓi and 231 inter-autosomal ℓij for the 1KG cohorts. There were ubiquitously existence of extended LD, which was associated with population structure or admixture (Figure 3).

We provided high-resolution illustration that decomposed a chromosome into upmost nearly a million grids, each of which was consisted of 250 X 250 SNP pairs, the highest resolution that has been realized so far at autosomal level (Figure 4); tremendous variable recombination rates led to regional strong LD as highlighted for the HLA region of chromosomes 6 and the centromere region of chromosome 11.

Furthermore, a consequently linear regression constructed could quantify LD decay score genome-widely, and in contrast LD decay was previously surrogated in a computational expensive method. There was strong ethnicity effect that was associated with extended LD (Figure 5).

We demonstrate that the strength of autosomal ℓi was inversely proportional to the SNP number, an anticipated relationship that is consistent to genome-wide spread of recombination hotspots. However, the chromosome 8 of ASW showed substantial deviation from the fitted linear relationship (Figure 6).

Schematic illustration for large-scale LD analysis as exampled for CONVERGE cohort.

A) The 22 human autosomes have consequently 22 and 231 , without (left) and with (right) scaling transformation; Scaling transformation is given in Eq 8. B) If zoom into chromosome 2 of 420,946 SNPs, a chromosome of relative neutrality is expected to have self-similarity structure that harbors many approximately strong along the diagonal, and relatively weak off-diagonally. Here chromosome 2 of CONVERGE has been split into 1,000 blocks and yielded 1000 LD grids, and 499,500 LD grids.

Reconciliation for LD estimators in the 26 1KG cohorts.

A) Consistency examination for the 26 1KG cohorts for their and estimated by X-LD and PLINK (--r2). In each figure, the 22 fitting line is in purple, whereas the 231 fitting line is in green. The gray solid line, , in which the sample size of each cohort, represents the expected fit between PLINK and X-LD estimates, and the two estimated regression models at the top-right corner of each plot shown this consistency. The sample size of each cohort is in parentheses. B) Distribution of R2 of and fitting lines is based on X-LD and PLINK algorithms in the 26 cohorts.

26 1KG cohorts: MSL (Mende in Sierra Leone), GWD (Gambian in Western Division, The Gambia), YRI (Yoruba in Ibadan, Nigeria), ESN (Esan in Nigeria), ACB (African Caribbean in Barbados), LWK (Luhya in Webuye, Kenya), ASW (African Ancestry in Southwest US); CHS (Han Chinese South), CDX (Chinese Dai in Xishuangbanna, China), KHV (Kinh in Ho Chi Minh City, Vietnam), CHB (Han Chinese in Beijing, China), JPT (Japanese in Tokyo, Japan); BEB (Bengali in Bangladesh), ITU (Indian Telugu in the UK), STU (Sri Lankan Tamil in the UK), PJL (Punjabi in Lahore, Pakistan), GIH (Gujarati Indian in Houston, TX); TSI (Toscani in Italia), IBS (Iberian populations in Spain), CEU (Utah residents (CEPH) with Northern and Western European ancestry), GBR (British in England and Scotland), FIN (Finnish in Finland); MXL (Mexican Ancestry in Los Angeles, California), PUR (Puerto Rican in Puerto Rico), CLM (Colombian in Medellin, Colombia), PEL (Peruvian in Lima, Peru).

Various LD components for the 26 1KG cohorts.

A) Chromosomal scale LD components for 5 representative cohorts (CEU, CHB, YRI, ASW, and 1KG). The upper parts of each figure represent (along the diagonal) and (off-diagonal), and the lower part as in Eq 8. For visualization purposes, the quantity of LD before scaling is transformed to a -log10 scale, with smaller values (red hues) representing larger LD, and a value of 0 representing that all SNPs are in LD. B) The relationship between the degree of population structure (approximated by ) and , and in the 26 1KG cohorts.

High-resolution illustration for LD grids for CEU, CHB, YRI, and ASW (m= 250).

For each cohort, we partition chromosomes 6 and 11 into high-resolution LD grids (each LD grid contains 250 × 250 SNP pairs). The bottom half of each figure shows the LD grids for the entire chromosome. Further zooming into HLA on chromosome 6 and the centromere region on chromosome 11, and their detailed LD in the relevant regions are also provided in the upper half of each figure. For visualization purposes, LD is transformed to a -log10-scale, with smaller values (red hues) representing larger LD, and a value of 0 representing that all SNPs are in LD.

LD decay analysis for 26 1KG cohorts.

A) Conventional LD decay analysis in PLINK for 26 cohorts. To eliminate the influence of sample size, the inverse of sample size has been subtracted from the original LD values. The YRI cohort, represented by the orange dotted line, is chosen as the reference cohort in each plot. The top-down arrow shows the order of LDdecay values according to Table 4. B) Model-based LD decay analysis for the 26 1KG cohorts. We regressed each autosomal against its corresponding inversion of the SNP number for each cohort. Regression coefficient quantifies the averaged LD decay of the genome and intercept provides a direct estimate of possible existence of long-distance LD. The values in the first three plot indicate the correlation between and LD decay score in three different physical distance and the correlation between (left-side vertical axis) and LD decay score (right-side vertical axis) and the correlation between (left-side vertical axis) and (right-side vertical axis), respectively. The last plot assessed the impact of centromere region of chromosome 11 on the linear relationship between chromosomal LD and the inverse of the SNP number. The dark and light gray dashed lines represent the mean of the ℛ with and without the presence of centromere region of chromosome 11.

The correlation between the inversion of the SNP number and .

A) The correlation between the inversion of the SNP number and in CEU, CHB, YRI, and ASW. B) Leave-one-chromosome-out strategy is adopted to evaluate the contribution of a certain chromosome on the correlation between the inverse of the SNP number and . C) The correlation between the inversion of the SNP number and chromosomal LD in CEU, CHB, YRI, and ASW after removing the centromere region of chromosome 11. D) High-resolution illustration for LD grids for chromosome 8 in CEU, CHB, YRI, and ASW. For each cohort, we partition chromosome 8 into consecutive LD grids (each LD grid contains 250 × 250 SNP pairs). For visualization purposes, LD is transformed to a -log10-scale, with smaller values (red hues) representing larger LD, and a value of 0 representing that all SNPs are in LD.

The proposed algorithm has been realized in C++ and is available at: https://github.com/gc5k/gear2. As tested the software could handle sample size as large as more than 10,000 individuals.

Methods and Materials

The overall rationale for large-scale LD analysis

We assume LD for a pair of biallelic loci is measured by the squared Pearson’s correlation, , in which the LD of loci l1 and l2, p and q the reference and the alternative allele frequencies. If we consider the averaged LD for a genomic grid over SNP pairs, the conventional estimator is , and, if we consider the averaged LD for mi and mj SNP pairs between two genomic segments, then . Now let us consider the 22 human autosomes (Figure 1A). We naturally partition the genome into 𝒞 = 22 blocks, and its genomic LD, denoted as ℓg, can be expressed as So we can decompose ℓg into 𝒞 ℓi and unique ℓij. Obviously, Eq 1 can be also expressed in the context for a single chromosome , in which the number of m SNP segments, each of which has SNPs. Geometrically it leads to ℬi diagonal grids and unique off-diagonal grids (Figure 1B).

LD-decay regression

As human genome can be boiled down to small LD blocks by genome-widely spread recombination hotspots (Hinch et al., 2019; Li et al., 2022), mechanically there is self-similarity for each chromosome that the relatively strong ℓi for juxtaposed grids along the diagonal but weak ℓi for grids slightly off-diagonal. So, for a chromosomal ℓi, we can further express it as in which ℓu is the mean LD for a diagonal grid, ℓuν the mean LD for off-diagonal grids, and mi the number of SNPs on the ith chromosome. Consider a linear model below, in which the inversion of the SNP number of the ith chromosome. After some algebra, if E(ℓu) ≫ E(ℓuV) – say if the former is one order greater than the latter, the interpretation of b1 and b0 can be It should be noticed that E(b1) ≈ (ℓu)m quantifies the averaged LD decay of the genome. Conventional LD decay is analysed via the well-known LD decay analysis, but Eq 4 provides a direct estimate of both LD decay and possible existence of extended LD. We will see the application of the model in Figure 5 that the strength of the long-distance LD is associated with population structure. Of note, the underlying assumption of Eq 3 and Eq 4 is genome-wide spread of recombination hotspots, an established result that has been revealed and confirmed (Hinch et al., 2019).

Efficient estimation for fg, fi, and fij

For the aforementioned analyses, the bottleneck obviously lies in the computational cost in estimating ℓi and ℓij. ℓi and ℓij are used to be estimated via the current benchmark algorithm as implemented in PLINK (Chang et al., 2015), and the computational time complex is proportional to 𝒪(nm2). We present a novel approach to estimate ℓi and ℓij. Given a genotypic matrix X, a n × m matrix, if we assume that there are mi and mj SNPs on chromosomes i and j, respectively, we can construct n × n genetic relatedness matrices as below in which is the standardized Xi and , where xkl is the genotype for the kth individual at the lth biallelic locus, F is the inbreeding coefficient having the value of 0 for random mating population and 1 for an inbred population, pl and ql are the frequencies of the reference and the alternative alleles (pl + ql = 1), respectively. When GRM is given, we can obtain some statistical characters of Gi. From Gi, we extract lower-triangle off-diagonal matrix and diagonal matrix , then we decompose . The mathematical expectation of , in which , can be established according to Isserlis’s theorem in terms of the four-order moment (Isserlis, 1918), in which is the expected relatedness score. r = 0 for the same individual, and r = 1 for first degree of relatives. Similarly, we can derive for . Eq 6 establishes the connection between GRM and the aggregated LD estimation that . According to Delta method (Lynch and Walsh, 1998), the means and the sampling variances for ℓi and ℓij are, in which and , respectively. Of note, the properties of ℓg can be derived similarly if we replace ℓi with ℓg in Eq 7. We can develop , a scaled version of ℓij, as below in which , a modification that removed the LD with itself. According to Delta method, the sampling variance of is Of note, when there is no LD between a pair of loci, ℓ yields zero and its counterpart PLINK estimate yields , a difference that can be reconciled in practice (see Figure 2).

Raise of LD due to population structure

In this study, the connection between LD and population structure is bridged via two pathways below, in terms of a pair of loci and of the aggregated LD for all pair of loci. For a pair of loci, their LD is often simplified as , but will be inflated if there are subgroups (Nei and Li, 1973). In addition, it is well established the connection between population structure and eigenvalues, and in particular the largest eigenvalue is associated with divergence of subgroups (Patterson et al., 2006). In this study, the existence of subgroups of cohort is surrogated by the largest eigenvalue λ1 or .

Data description and quality control

The 1KG (Auton et al., 2015), which is launched to produce a deep catalogue of human genomic variation by whole genome sequencing (WGS) or whole exome sequencing (WES), and 2,503 strategically selected individuals of global diversity are included (containing 26 cohorts). We used the following criteria for SNP inclusion for each of the 26 1KG cohorts: i) autosomal SNPs only; ii) SNPs with missing genotype rates higher than 0.2 were removed, and missing genotypes were imputed; iii) Only SNPs with minor allele frequencies higher than 0.05 were retained. Then 2,997,635 consensus SNPs that were present in each of the 26 cohorts were retained. According to their origins, the 26 cohorts are grouped as African (AFR: MSL, GWD, YRI, ESN, ACB, LWK, and ASW), European (EUR: TSI, IBS, CEU, GBR, and FIN), East Asian (EA: CHS, CDX, KHV, CHB, and JPT), South Asian (SA: BEB, ITU, STU, PJL, and GIH), and American (AMR: MXL, PUR, CLM, and PEL), respectively.

In addition, to test the capacity of the developed software (X-LD), we also included CONVERGE cohort (n = 10,640), which was used to investigate Major Depressive Disorder (MDD) in the Han Chinese population (Cai et al., 2015). We performed the same criteria for SNP inclusion as that of the 1KG cohorts, and m = 5,215,820 SNPs were remained for analyses.

X-LD software implementation

The proposed algorithm has been realized in our X-LD software, which X-LD is written in C++ and reads in binary genotype data as often used in PLINK. As multi-thread programming is adopted, the efficiency of X-LD can be improved upon the availability of computational resources. We have tested X-LD in various independent datasets for its reliability and robustness. Certain data management options, such as flexible inclusion or exclusion of chromosomes, have been built into the commands of X-LD. In X-LD, missing genotypes are naively imputed according to Hardy-Weinberg proportions.

The most time-consuming part of X-LD was the construction of GRM , and the established computational time complex was 𝒪(n2m). However, if is decomposed into , in which has dimension of n × B, using Mailman algorithm the computational time complex for building G can be reduced to (Liberty and Zucker, 2009). This idea of embedding Mailman algorithm into certain high throughput genomic studies has been successful, and our X-LD software is also leveraged by absorbing its recent practice (Wu and Sankararaman, 2018).

Results

Statistical properties of the proposed method

As schematically illustrated in Figure 1, ℓg could be decomposed into 𝒞 ℓi and unique ℓij components. We compared the estimated ℓi and ℓi in X-LD with those being estimated in PLINK (known as “--r2”). Considering the substantial computational cost of PLINK, only 100,000 randomly selected autosome SNPs were used for each 1KG cohort, and 22 and 231 were estimated. After regressing 22 against those of PLINK, we found that the regression slope was close to unity and bore an anticipated intercept a quantity of approximately (Figure 2A and Figure 2B). In other words, PLINK gave even for SNPs of no LD. However, when regressing 231 estimates against those of PLINK, it was found that largely because of tiny quantity of it was slightly smaller than 1 but statistically insignificant from 1 in these 26 1KG cohorts (mean of 0.86 and s.d. of 0.10, and its 95 % confidence interval was (0.664, 1.056)); when the entire 1KG samples were used, its much larger LD due to subgroups, nearly no estimation bias was found (Figure 2A and Figure 2B). In contrast, because of their much larger values, components were always consistent with their corresponding estimates from PLINK (mean of 1.03 and s.d. of 0.012, 95% confidence interval was (1.006, 1.053), bearing an ignorable bias). Furthermore, we also combined the African cohorts together (MSL, GWD, YRI, ESN, LWK, totaling 599 individuals), the East Asian cohorts together (CHS, CDX, KHV, CHB, and JPT, totaling 504 individuals), and the European cohorts together (EUR: TSI, IBS, CEU, GBR, and FIN, totaling 503 individuals), the resemblance pattern between X-LD and PLINK was similar as observed in each cohort alone (Figure S1). The empirical data in 1KG verified that the proposed method was sufficiently accurate.

To fairly evaluate the computational efficiency of our proposed method, the benchmark comparison was conducted on the first chromosome of the entire 1KG dataset (n = 2,503 and m = 225,967), and 10 CPUs were used for multi-thread computing. Compared with PLINK, the calculation efficiency of X-LD was nearly 30∼40 times faster for the tested chromosome, and its computational time of X-LD was proportional to (Figure S2). So, X-LD provided a feasible and reliable estimation of large-scale complex LD patterns. More detailed computational time of the tested tasks would be reported in their corresponding sections below; since each 1KG cohort has sample size around 100, otherwise specified the computational time was only reported for CHB (n = 103) as a reference (Table 1). In order to test the capability of the software, the largest dataset tested was CONVERGE (n = 10,640, and m = 5,215,820), and it took 77,508.00 seconds, about 22 hours, to estimate 22 autosomal and 231 (Figure 1A); When zooming into chromosome 2 of CONVERGE, on which 420,949 SNP had been evenly split into 1,000 blocks and yielded 1000 , grids, and 499,500 LD grids, it took 45,125.00 seconds, about 12.6 hours, to finished the task (Figure 1B).

Ubiquitously extended LD and population structure/admixture

We partitioned the 2,997,635 SNPs into 22 autosomes (Figure 3A and Figure S3), and the general LD patterns were as illustrated for CEU, CHB, YRI, ASW, and 1KG. As expected, for each cohort (Figure 3B). As observed in these 1KG cohorts, all these three LD measures were associated with population structure, which was surrogated by , and their squared correlation R2 were greater than 0.8. ACB, ASW, PEL, and MXL, which all showed certain admixture, tended to have much greater , and (Table 2 and Figure 3B). In contrast, East Asian (EA) and European (EUR) orientated cohorts, which showed little within cohort genetic differentiation – as their largest eigenvalues were slightly greater than 1, had their aggregated LD relatively low and resembled each other (Table 2). Furthermore, for several European (TSI, IBS, and FIN) and East Asian (JPT) cohorts, the ratio between and components could be smaller than 0.1, and the smallest ratio was found to be about 0.091 in FIN. The largest ratio was found in 1KG that and , and the ratio was 0.877 because of the inflated LD due to population structure. A more concise statistic to describe the ratio between ℓij and ℓi was , Eq 8, and the corresponding values for 231 scaled for FIN was (s.d. of 0.027) and for 1KG was (s.d. of 0.028).

In terms of computational time, for 103 CHB samples, it took about 101.34 seconds to estimate 22 autosomal and 231 ; for all 1KG 2,503 samples, X-LD took about 3,008.29 seconds (Table 1). Conventional methods took too long to complete the analyses in this section, so no comparable computational time was provided. For detailed 22 and 231 estimates for each 1KG cohort, please refer to Extended Data 1 (Excel Sheet 1-27).

Detecting exceedingly high LD grids shaped by variable recombination rates

We further explored each autosome with high-resolution grid LD visualization. We set m = 250, so each grid had the ℓuv for 250 × 250 SNP pairs. The computational time complex was , in which , and with our proposed method in CHB it costed 66.86 seconds for chromosome 2, which had the most 241,241 SNPs and was totaled 466,095 unique grids, and 3.22 seconds for chromosome 22, which had the least 40,378 SNPs and was totaled 13,203 unique grids (Table 1). In contrast, under conventional methods those LD grids were not very likely to be exhaustively surveyed because of its computational cost was : for CHB chromosome 2, it would have taken about 40 hours as estimated. As the result was very similar for m = 250 (Figure S4), we only reported the results under m = 250 below.

As expected, chromosome 6 (206,165 SNPs, totaling 340,725 unique grids) had its HLA cluster showing much higher LD than the rest of chromosome 6. In addition, we found very dramatic variation of HLA cluster LD (28,477,797-33,448,354 bp, totaling 3,160 unique grids) across ethnicities. For CEU, CHB, YRI, and ASW, their , and 0.0019, respectively, but their corresponding HLA cluster grids had , and 0.022, respectively (Figure 4). Consequently, the largest ratio for was of 42.00 in CEU, 39.06 in YRI, and 32.22 in CHB, but was reduced to 11.58 in ASW. Before the release of CHM13 (Hoyt et al., 2022), chromosome 11 had the most completely sequenced centromere region, which had much rarer recombination events, all four cohorts showed an strong LD around the centromere (46,061,947-59,413,484 bp, totaling 1,035 unique grids) regardless of their ethnicities (Figure 4). , and 0.0022, respectively, and , and 0.094, respectively; the ratio for , and 94,05, for CEU, CHB, and YRI, respectively; the lowest ratio was found in ASW of 42.73. In addition, removing the HLA region of chromosome 6 or the centromere region of chromosome 11 would significantly reduce or in comparison with the randomly removal of other regions (Figure S5).

Model-based LD decay regression revealed LD composition

The real LD block size was not exact of m = 250 or m = 500, but an unknown parameter that should be inferred in computational intensive “LD decay” analysis (Zhang et al., 2019; Chang et al., 2015). We conducted the conventional LD decay for the 26 1KG cohorts (Figure 5A), and the time cost was 1,491.94 seconds for CHB. For each cohort, we took the area under the LD decay curve in the LD decay plot, and it quantified approximately the LD decay score for each cohort. The smallest score was 0.0421 for MSL and the largest was 0.0598 for PEL (Table 4). However, this estimation was not taken into account the real extent of LD, so it was not precise enough to reflect the LD decay score. For example, for admixture population, such as American cohorts, the extent of LD would be longer.

In contrast, we proposed a model-based method, as given in Eq 3, which could estimate LD decay score (regression coefficient b1) and long-distance LD score (intercept b0) jointly. Given the estimated 22 (Extended data 1; Table 3 for four representative cohorts), we regressed each autosomal against its correspondingly inversion of SNP number, and all yielded positive slopes (Pearson’s correlation ℛ > 0.80, Table 4; Figure 5B), an observation that was consistent with genome-wide spread of recombination hotspots. This linear relationship could consequently be considered the norm for a relative homogenous population as observed in most 1KG cohorts (Figure S6), while for the all 2,503 1KG samples ℛ = 0.55 only (Table 4), indicating that the population structure and possible differentiated recombination hotspots across ethnicities disturbed the assumption underlying Eq 3 and smeared the linearity. We extracted and for the 26 1KG cohorts for further analysis. The rates of LD decay score, as indicated by , within the African cohorts (AFR) were significantly faster than other continents, consistent with previous observation that African population had relative shorter LD (Gabriel et al., 2002); while subgroups within the American continent (AMR) tended to have extended LD range due to their admixed genetic composition (Table 4 and Figure 5B). Notably, the correlation between and the approximated LD decay score was ℛ = 0.88. The estimated were highly correlated with (ℛ = 0.94).

A common feature was universally relative high LD of chromosome 6 and 11 in the 26 1KG cohorts (Figure S6). We quantified the impact of chromosome 6 and 11 by leave-one-chromosome-out test in CEU, CHB, YRI, and ASW for details (Figure 6A and 6B), and found that chromosome 6 could lift ℛ on average by 0.017, and chromosome 11 by 0.046. One possible explanation was that the centromere regions of chromosomes 6 and 11 have been assembled more completely than other chromosomes before the completion of CHM13 (Hoyt et al., 2022), whereas meiotic recombination tended to be reduced around the centromeres (Hinch et al., 2019). We estimated ℓi after having knocked out the centromere region (46,061,947-59,413,484 bp, chr 11) in CEU, CHB, YRI, and ASW, and chromosome 11 then did not deviate much from their respective fitted lines (Figure 6C). A notable exceptional pattern was found in ASW, the chromosome 8 of which had even more deviation than chromosome 11 (ℛ was 0.83 and 0.87 with and without chromosome 8 in leave-one-chromosome out test) (Figure 6B). The deviation of chromosome 8 of ASW was consistent even more SNPs were added (Figure S7). We also provided high-resolution LD grids illustration for chromosome 8 (163,436 SNPs, totaling 214,185 grids) of the four representative cohorts for more detailed virtualization (Figure 6D). ASW had , but 0.00075, 0.00069 and 0.00043 for CEU, CHB, and YRI, respectively.

Discussion

In this study, we present a computationally efficient method to estimate mean LD of genomic grids of many SNP pairs. Our LD analysis framework is based on GRM, which has been embedded in variance component analysis for complex traits and genomic selection (Goddard, 2009; Visscher et al., 2014; Chen, 2014). The key connection from GRM to LD was bridged via the transformation between n × n matrix and m × m matrix, in particular here via Isserlis’s theorem under the fourth-order moment (Isserlis, 1918). With this connection, the computational cost for estimating the mean LD of m × m SNP pairs is reduced from 𝒪 (n2m) to 𝒪 (n2m), and the statistical properties of the proposed method are derived in theory and validated in 1KG datasets. In addition, as the genotype matrix X is of limited entries {0, 1, 2}, assuming missing genotypes are imputed first, using Mailman algorithm the computational cost of GRM can be further reduced to (Liberty and Zucker, 2009). The largest data tested so far for the proposed method has the sample size of 10,640 and of more than 5 million SNPs, it can complete genomic LD analysis in 77,508.00 seconds (Table 1). Obviously, with the availability of such as UK Biobank data (Bycroft et al., 2018), the proposed method may not be adequate and other new methods are needed.

We also applied the proposed method into 1KG and revealed certain characteristics of the human genomes. Firstly, we found the ubiquitously existence of extended LD, which was likely emerged because of population structure, even very slightly, and admixture history. We quantified the and in 1KG, and as indicated by we found the inter-chromosomal LD was nearly an order lower than intra-chromosomal LD; for admixed cohorts, the ratio was much higher, even very close to each other such as in all 1KG samples. Secondly, variable recombination rates shaped peak of local LD. For example, the HLA region showed high LD in European and East Asian cohorts, but relatively low LD in such as YRI, consistent with their much longer population history. Thirdly, it existed general linear correlation between ℓi and the inversion of the SNP number, a long-anticipated result that is as predicted with genome-wide spread of recombination hotspots (Hinch et al., 2019). One outlier of this linear norm was chromosome 11, which had so far most completely genotyped centromere and consequently had more elevated LD compared with other autosomes. We anticipate that with the release of CHM13 the linear correlation should be much closer to unity (Hoyt et al., 2022). Of note, under the variance component analysis for complex traits, it is often a positive correlation between the length of a chromosome (as surrogated by the number of SNPs) and the proportion of heritability explained (Chen et al., 2014).

In contrast, throughout the study recurrent outstanding observations were found in ASW. For example, in ASW the ratio of was substantially dropped down compared with that of CEU, CHB, or YRI as illustrated in Figure 4. Furthermore, chromosome 8 in ASW fluctuated upwards most from the linear correlation (Figure 6), and even after various analyses, such as expanding SNP numbers. One possible explanation may lay under the complex demographic history of ASW, which can be investigated and tested in additional African American samples or possible existence for epistatic fitness (Ni et al., 2020).

Supporting information

eLife Assessment

This study presents a valuable new approach for efficient computation of statistics on correlations between genetic variants (linkage disequilibrium, or LD), which the authors apply to quantify the extent of LD across chromosomes. The method appears solid, although the presentation of equations needs clarification and improvement. The authors document that cross-chromosome LD can be substantial, which has implications for geneticists who are interested in population structure and its impact on genetic association studies.

Reviewer #1 (Public Review):

Summary:

Huang and colleagues present a method for approximation of linkage disequilibrium (LD) matrices. The problem of computing LD matrices is the problem of computing a correlation matrix. In the cases considered by the authors, the number of rows (n), corresponding to individuals, is small compared to the number of columns (m), corresponding to the number of variants. Computing the correlation matrix has cubic time complexity [O(nm^2)], which is prohibitive for large samples. The authors approach this using three main strategies: 1. they compute a coarsened approximation of the LD matrix by dividing the genome into variant-wise blocks which statistics are effectively averaged over; 2. they use a trick to get the coarsened LD matrix from a coarsened genomic relatedness matrix (GRM), which, with O(n^2 m) time complexity, is faster when n << m; 3. they use the Mailman algorithm to improve the speed of basic linear algebra operations by a factor of log(max(m,n)). The authors apply this approach to several datasets.

Strengths:

- the authors demonstrate that their proposed method performs in line with theoretical explanations

- the coarsened LD matrix is useful for describing global patterns of LD, which do not necessarily require variant-level resolution

- they provide an open-source implementation of their software

Weaknesses:

- the coarsened LD matrix is of limited utility outside of analyzing macroscale LD characteristics

- the method still essentially has cubic complexity--albeit the factors are smaller and Mailman reduces this appreciably. It would be interesting if the authors were able to apply randomized or iterative approaches to achieve more fundamental gains. The algorithm remains slow when n is large and/or the grid resolution is increased.

Reviewer #2 (Public Review):

Summary:

In this paper, the authors point out that the standard approach of estimating LD is inefficient for datasets with large numbers of SNPs, with a computational cost of O(nm^2), where n is the number of individuals and m is the number of SNPs. Using the known relationship between the LD matrix and the genomic-relatedness matrix, they can calculate the mean level of LD within the genome or across genomic segments with a computational cost of O(n^2m). Since in most datasets, n<

Strengths:

Generally, for computational papers like this, the proof is in the pudding, and the authors appear to have been successful at their aim of producing an efficient computational tool. The most compelling evidence of this in the paper is Figure 2 and Supplementary Figure S2. In Figure 2, they report how well their X-LD estimates of LD compare to estimates based on the standard approach using PLINK. They appear to have very good agreement. In Figure S2, they report the computational runtime of X-LD vs PLINK, and as expected X-LD is faster than PLINK as long as it is evaluating LD for more than 8000 SNPs.

Weakness:

While the X-LD software appears to work well, I had a hard time following the manuscript enough to make a very good assessment of the work. This is partly because many parameters used are not defined clearly or at all in some cases. My best effort to intuit what the parameters meant often led me to find what appeared to be errors in their derivation. As a result, I am left worrying if the performance of X-LD is due to errors cancelling out in the particular setting they consider, making it potentially prone to errors when taken to different contexts.

Impact:

I feel like there is value in the work that has been done here if there were more clarity in the writing. Currently, LD calculations are a costly step in tools like LD score regression and Bayesian prediction algorithms, so a more efficient way to conduct these calculations would be useful broadly. However, given the difficulty I had following the manuscript, I was not able to assess when the authors' approach would be appropriate for an extension such as that.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review)

Summary:

Huang and colleagues present a method for approximation of linkage disequilibrium (LD) matrices. The problem of computing LD matrices is the problem of computing a correlation matrix. In the cases considered by the authors, the number of rows (n), corresponding to individuals, is small compared to the number of columns (m), corresponding to the number of variants. Computing the correlation matrix has cubic time complexity , which is prohibitive for large samples. The authors approach this using three main strategies:

1. they compute a coarsened approximation of the LD matrix by dividing the genome into variant-wise blocks which statistics are effectively averaged over;

1. they use a trick to get the coarsened LD matrix from a coarsened genomic relatedness matrix (GRM), which, with time complexity, is faster when n << m;

1. they use the Mailman algorithm to improve the speed of basic linear algebra operations by a factor of log(max(m,n)). The authors apply this approach to several datasets.

Strengths:

The authors demonstrate that their proposed method performs in line with theoretical explanations.

The coarsened LD matrix is useful for describing global patterns of LD, which do not necessarily require variant-level resolution.

They provide an open-source implementation of their software.

Weaknesses:

The coarsened LD matrix is of limited utility outside of analyzing macroscale LD characteristics. The method still essentially has cubic complexity--albeit the factors are smaller and Mailman reduces this appreciably. It would be interesting if the authors were able to apply randomized or iterative approaches to achieve more fundamental gains. The algorithm remains slow when n is large and/or the grid resolution is increased.

Thanks for your positive and accurate evaluation! We acknowledge the weakness and include some sentences in Discussion.

“The weakness of the proposed method is obvious that the algorithm remains slow when the sample size is large or the grid resolution is increased. With the availability of such as UK Biobank data (Bycroft et al., 2018), the proposed method may not be adequate, and much advanced methods, such as randomized implementation for the proposed methods, are needed.”

Reviewer #2 (Public Review)

Summary:

In this paper, the authors point out that the standard approach of estimating LD is inefficient for datasets with large numbers of SNPs, with a computational cost of , where n is the number of individuals and m is the number of SNPs. Using the known relationship between the LD matrix and the genomic- relatedness matrix, they can calculate the mean level of LD within the genome or across genomic segments with a computational cost of . Since in most datasets, n<<m, this can lead to major computational improvements. They have produced software written in C++ to implement this algorithm, which they call X-LD. Using the output of their method, they estimate the LD decay and the mean extended LD for various subpopulations from the 1000 Genomes Project data.

Strengths:

Generally, for computational papers like this, the proof is in the pudding, and the authors appear to have been successful at their aim of producing an efficient computational tool. The most compelling evidence of this in the paper is Figure 2 and Supplementary Figure S2. In Figure 2, they report how well their X- LD estimates of LD compare to estimates based on the standard approach using PLINK. They appear to have very good agreement. In Figure S2, they report the computational runtime of X-LD vs PLINK, and as expected X-LD is faster than PLINK as long as it is evaluating LD for more than 8000 SNPs.

Weakness:

While the X-LD software appears to work well, I had a hard time following the manuscript enough to make a very good assessment of the work. This is partly because many parameters used are not defined clearly or at all in some cases. My best effort to intuit what the parameters meant often led me to find what appeared to be errors in their derivation. As a result, I am left worrying if the performance of X-LD is due to errors cancelling out in the particular setting they consider, making it potentially prone to errors when taken to different contexts.

Thanks for you critical reading and evaluation. We do feel apologize for typos, which have been corrected and clearly defined now (see Eq 1 and Table 1). In addition, we include more detailed mathematical steps, which explain how LD decay regression is constructed and consequently finds its interpretation (see the detailed derivation steps between Eq 3 and Eq 4).

Impact:

I feel like there is value in the work that has been done here if there were more clarity in the writing. Currently, LD calculations are a costly step in tools like LD score regression and Bayesian prediction algorithms, so a more efficient way to conduct these calculations would be useful broadly. However, given the difficulty I had following the manuscript, I was not able to assess when the authors’ approach would be appropriate for an extension such as that.

See our replies below in responding to your more detailed questions.

Reviewer #1 (Recommendations For The Authors)

There are numerous linguistic errors throughout, making it challenging to read.

It is unclear how the intercepts were chosen in Figure S2. Since theory only gives you the slopes, it seems like it would make more sense to choose the intercept such that it aligns with the empirical results in some way.

Thanks for your critical evaluation. We do feel apologize some typos, and we have read it through and clarify the text as much as possible. In addition, we included Table 1, which introduces mathematical symbols of the paper.

In Figure S2, the two algorithms being compared have different software implementations, PLINK vs X-LD. Their real performance not only depended on the time complexity of the algorithms (right-side y-axis), but also how the software was coded. PLINK is known for its excellent programming. If we could have programmed as well as Chris Chang, the performance of X-LD should have been even better and approach the ratio m/n. However, even under less skilled programming, X-LD outperformed plink.

Reviewer #2 (Recommendations For The Authors):

Thank you for the chance to review your manuscript. It looks like compelling work that could be improved by greater detail. Providing the level of detail necessary may require creating a Supplementary Note that does a lot of hand-holding for readers like me who are mathematically literate but who don’t have the background that you do. Then you can refer readers to the Supplement if they can’t follow your work.

We fix the problems and style issues as possible as we can.

Regarding the weakness section in the public review, here are a few examples of where I got confused, though this list is not exhaustive.

1. Consider Equation 1 (line 100), which I believe must be incorrect. Imagine that g consists of two SNPs on different chromosomes with correlation rho. Then ell_g (which is defined as the average squared elements of the correlation matrix) would be

ell_g = 1/4 (1 + 1 + rho^2 + rho^2) = (1+rho^2)/2.

But ell_1=1 and ell_2=1 and ell_12=rho^2 (The average squared elements of the chromosome-specific correlation matrices and the cross-chromosome correlation matrix, respectively). So

sum(ell_i)+sum(ell_ij) = 1 + 1 + rho^2 + rho^2 = (1+rho^2)*2.

I believe your formulas would hold if you defined your LD values as the sum of squared correlations instead of the mean, but then I don’t know if the math in the subsequent sections holds. I think this problem also holds for Eq 2 and therefore makes Eqs 3 and 4 difficult to interpret.

Thanks for your attentive review and invaluable suggestions. We acknowledge the typo in calculating the mean in Eq 1, resulting in difficulties in understanding the equations. We sincerely apologize for this oversight. To address this issue and ensure clarity in the interpretation of Eq 3 and Eq 4, we have provided more detailed explanations (see the derivation between Eq 3 and Eq 4).

1. I didn’t know what the parameters are in Equation 3. The vector ell needs to be defined. Is it the vector of ell_i for each chromosomal segment i? I’m also confused by the definition of m_i, which is defined on line 113 as the “SNP number of the i-th chromosome.” Do the authors mean the number of SNPs on the i-th chromosomal segment? If so, it wasn’t clear to me how Eq 2 and Eq 3 imply Eq 4. Further, it wasn’t clear to me why E(b1) quantifies the average LD decay of the genome. I’m used to seeing plots of average LD as a function of distance between SNPs to calculate this, though I’m admittedly not a population geneticist, so maybe this is standard. Standard or not, readers deserve to have their hands held a bit more through this either in the text or in a Supplementary Note.

Thanks for your insightful feedback. When we were writing this paper, our actually focus was Eq 3 and to establish the relationship between chromosomal LD and the reciprocal of the length of chromosome (Fig 6A) – which was surrogated by the number of SNPs, the correlation between ell_i and 1/m_i.

We asked around our friends who are population geneticists, who anticipated the correlation between chromosomal LD (ell) and 1/m. The rationale simple if one knows the very basis of population genetics. A long chromosome experiences more recombination, which weakens LD for a pair of loci. In particular, for a pair of loci D_t=D_0 (1-c)^t. D_t the LD at the t generation, D_0 at the 0 generation, and c the recombination fraction. As recombination hotspots are nearly even distributed along the genome, such as reported by Science 2019;363:eaau8861, the chromosome will be broken into the shape in Author response image 1 (Fig 1C, newly added). Along the diagonal you see tight LD block, which will be vanished in the further as predicted by D_t equation, and any loci far away from each other will not be in LD otherwise raised by such as population structure. Ideally, we assume the diagonal block of aveage size of m×m and average LD of a SNP with other SNPs inside the diagonal block (red) is l_u; and, in contrast, off-diagonal average LD (light red) to be l_uv. This logic is hidden but employed in such as ld score regression and prs refinement using LD structure.

But, how to estimate chromosomal LD (ell), which is overwhelming as our friends said! So, the Figure 6A is logically anticipated by a seasoned population geneticist, but has never been realized because of is nightmare. Often, those signature patterns should have been employed as showcases in releasing new reference data, such as HapMap. However, to our knowledge, this signature linear relationship has never been illustrated in those reference data.

If you further test a population geneticist, if any chromosome will deviate from this line (Fig 6A)? The answer most likely will be chromosome 6 because of the LD tight HLA region. However, it is chromosome 11 because of its most completed sequenced centromere. Chr 11 is a surprise! With T2T sequenced population, Chr 11 will not deviate much. We predict!

However, we suspect whether people appreciate this point, we shift our focus to efficient computation of LD—which is more likely understood. We acknowledge the lack of clarity in notation definitions and the absence of the derivation for the interpretation of b1 and b0 for LD decay regression. So, we have added a table to provide an explanation of the notation (see the Table 1) and provided additional derivations, which explained how LD decay regression was derived (see the derivation between Eq 3 and Eq 4). Figure 1C provides illustration for the underlying assumption under LD.

The technique to bridge Eq 2~3 to Eq 4 is called “building interpretation”. It once was one of the kernel tasks for population genetics or statistical genetics, and a classical example is Haseman-Elston regression (Behavior Genetics, 1972, 2:3-19). When it is moving towards a data-driven style, the culture becomes “shut up, calculate”. Finding interpretation for a regression is a vanishing craftmanship, and people often end up with unclear results!

1. In line 135, it’s not clear to me what is meant by . If it is , then wouldn’t the resulting matrix be a matrix of zeros since is zero everywhere except the lower off-diagonal? So maybe it is ? But then later in that line, you say that the square of this matrix is the sum of several terms of the form . Are these the scalar elements of the G matrix? But then the sum is a scalar, which can’t be true since is a matrix.

Thanks for your attentive review. We indeed confused the definition of matrices and their elements, and should refer to the stacked off-diagonal elements of matrix . So, is a vector for variable – the relationship between sample i and j. We assume the reviewer use R software, then corresponds to mean .

See the text between Eq 5 and Eq 6.

“We extract two vectors , which stacks the off-diagonal elements of , and , which takes the diagonal elements of .”

In addition, , so the ground truth is that , but not zero.

To clarify these math symbols, we replace G with K, so as to be consistent with our other works (see Table 1).

To derive the means and the sampling variances for and , the Eq 7 can be established by some modifications on the Delta method as exampled in Appendix I of Lynch and Walsh’s book (Lynch and Walsh, 1998). We added this sentence near Eq 7 in the main text.

Version 2

Abstract

Abstract

In this study, we proposed an efficient algorithm (X-LD) for estimating LD patterns for a genomic grid, which can be of inter-chromosomal scale or of small segments. Compared with conventional methods, the proposed method was significantly faster, dropped from 𝒪 (nm2) to 𝒪 (n2m)—n the sample size and m the number of SNPs, and consequently we were permitted to explore in depth unknown or reveal long-anticipated LD features of the human genome. Having applied the algorithm for 1000 Genome Project (1KG), we found: I) The extended LD, driven by population structure, was universally existed, and the strength of inter-chromosomal LD was about 10% of their respective intra-chromosomal LD in relatively homogeneous cohorts, such as FIN and to nearly 56% in admixed cohort, such as ASW. II) After splitting each chromosome into upmost more than a half million grids, we elucidated the LD of the HLA region was nearly 42 folders higher than chromosome 6 in CEU and 11.58 in ASW; on chromosome 11, we observed that the LD of its centromere was nearly 94.05 folders higher than chromosome 11 in YRI and 42.73 in ASW. III) We uncovered the long-anticipated inversely proportional linear relationship between the length of a chromosome and the strength of chromosomal LD, and their Pearson’s correlation was on average over 0.80 for 26 1KG cohorts. However, this linear norm was so far perturbed by chromosome 11 given its more completely sequenced centromere region. Uniquely chromosome 8 of ASW was found most deviated from the linear norm than any other autosomes. The proposed algorithm has been realized in C++ (called X-LD) and available at https://github.com/gc5k/gear2, and can be applied to explore LD features in any sequenced populations.

Introduction

Linkage disequilibrium (LD) is the association for a pair of loci and the metric of LD serves as the basis for developing genetic applications in agriculture, evolutionary biology, and biomedical researches (Weir, 2008; Hill and Robertson, 1966). The structure of LD of the human genome is shaped by many factors, mutation, recombination, population demography, epistatic fitness, and completeness of genomic data itself (Myers et al., 2005; Nei and Li, 1973; Ardlie et al., 2002). Due to its overwhelming cost, LD structure investigation is often compromised to a small genomic region (Chang et al., 2015; Theodoris et al., 2021), and their typical LD structure is as illustrated for a small segment (Barrett et al., 2005). Now, given the availability of large-scale genomic data, such as millions of single nucleotide polymorphisms (SNPs), the large-scale LD patterns of the human genome play crucial roles in determining genomics studies, and many theories and useful algorithms upon large-scale LD structure, from genome-wide association studies, polygenic risk prediction for complex diseases, and choice for reference panels for genotype imputation (Vilhjálmsson et al., 2015; Yang and Zhou, 2020; Bulik-Sullivan et al., 2015; Yang et al., 2011; Das et al., 2016).

However, there are impediments, largely due to intensified computational cost, in both investigating large-scale LD and providing high-resolution illustration for their details. If we consider a genomic grid that is consisted of m2 SNP pairs, given a sample of n individuals and m SNPs (n «m) – typically as observed in 1000 Genomes Project (1KG) (Lowy-Gallego et al., 2019), its benchmark computational time cost for calculating all pairwise LD is 𝒪 (nm2), a burden that quickly drains computational resources given the volume of the genomic data. In practice, it is of interest to know the mean LD of the SNP pairs for a genomic grid, which covers mi ×mj SNP pairs. Upon how a genomic grid is defined, a genomic grid consequently can be consisted of : i) the whole genome-wide m2 SNP pairs, and we denote their mean LD as ℓg ; ii) the intra-chromosomal mean LD for the ith chromosome of SNP pairs, and denote as ℓi ; iii) the inter-chromosomal mean LD and chromosomal mi mj SNP pairs, and denoted as ℓi· j .

In this study we propose an efficient algorithm that can estimate ℓg, ℓi, and ℓ i· j ·, the computational time of which can be reduced from to 𝒪 (n2mi) for ℓi and 𝒪 (nmi mj) to 𝒪 (n2mi+n2mj ) for ℓi· j. The rationale of the proposed method relies on the connection between the genetic relationship matrix (GRM) and LD (Chen, 2014; Goddard, 2009), and in this study a more general transformation from GRM to LD can be established via Isserlis’s theorem (Isserlis, 1918; Zhou, 2017). The statistical properties, such as sampling variance, of the estimated LD have been derived too.

The proposed method can be analogously considered a more powerful realization for Haploview (Barrett et al., 2005), but additional utility can be derived to bring out unprecedented survey of LD patterns of the human genome. As demonstrated in 1KG, we consequently investigate how biological factors such as population structure, admixture, or variable local recombination rates can shape large-scale LD patterns of the human genomes.

The proposed method provides statistically unbiased estimates for large-scale LD patterns and shows computational merits compared with the conventional methods (Figure 2).

We estimated ℓg, and 22 autosomal ℓi. and 231 inter-autosomal ℓi· j for the 1KG cohorts. There were ubiquitously existence of extended LD, which was associated with population structure or admixture (Figure 3).

We provided high-resolution illustration that decomposed a chromosome into upmost nearly a million grids, each of which was consisted of 250×250 SNP pairs, the highest resolution that has been realized so far at autosomal level (Figure 4); tremendous variable recombination rates led to regional strong LD as highlighted for the HLA region of chromosomes 6 and the centromere region of chromosome 11.

Furthermore, a consequently linear regression constructed could quantify LD decay score genome-widely, and in contrast LD decay was previously surrogated in a computational expensive method. There was strong ethnicity effect that was associated with extended LD (Figure 5).

We demonstrate that the strength of autosomal ℓi. was inversely proportional to the SNP number, an anticipated relationship that is consistent to genome-wide spread of recombination hotspots. However, the chromosome 8 of ASW showed substantial deviation from the fitted linear relationship (Figure 6).

Schematic illustration for large-scale LD analysis as exampled for CONVERGE cohort.

A) The 22 human autosomes have consequently 22 and 231 , ut (left) and with (right) scaling transformation; Scaling transformation is given in Eq 8. B) If zoom into chromosome 2 of 420,946 SNPs, a chromosome of relative lity is expected to have self-similarity structure that harbors many approximately strong along the diagonal, and relatively weak off-diagonally. Here osome 2 of CONVERGE has been split into 1,000 blocks and yielded 1000 LD grids, and 499,500 LD grids. C) An illustration of the construction process for D-decay regression model

Reconciliation for LD estimators in the 26 cohorts of 1KG.

A) Consistency examination for the 26 1KG cohorts for their and estimated by X-LD and PLINK (--r2). In each figure, the 22 fitting line is in purple, whereas the 231 fitting line is in green. The gray solid line,, in which n the sample size of each cohort, represents the expected fit between PLINK and X-LD estimates, and the two estimated regression models at the top-right corner of each plot shown this consistency. The sample size of each cohort is in parentheses. B) Distribution of R2 of and fitting lines is based on X-LD and PLINK algorithms in the 26 cohorts; R2 represents variation explained by the fitted model. 26 1KG cohorts: MSL (Mende in Sierra Leone), GWD (Gambian in Western Division, The Gambia), YRI (Yoruba in Ibadan, Nigeria), ESN (Esan in Nigeria), ACB (African Caribbean in Barbados), LWK (Luhya in Webuye, Kenya), ASW (African Ancestry in Southwest US); CHS (Han Chinese South), CDX (Chinese Dai in Xishuangbanna, China), KHV (Kinh in Ho Chi Minh City, Vietnam), CHB (Han Chinese in Beijing, China), JPT (Japanese in Tokyo, Japan); BEB (Bengali in Bangladesh), ITU (Indian Telugu in the UK), STU (Sri Lankan Tamil in the UK), PJL (Punjabi in Lahore, Pakistan), GIH (Gujarati Indian in Houston, TX); TSI (Toscani in Italia), IBS (Iberian populations in Spain), CEU (Utah residents (CEPH) with Northern and Western European ancestry), GBR (British in England and Scotland), FIN (Finnish in Finland); MXL (Mexican Ancestry in Los Angeles, California), PUR (Puerto Rican in Puerto Rico), CLM (Colombian in Medellin, Colombia), PEL (Peruvian in Lima, Peru).

Various LD components for the 26 1KG cohorts.

A) Chromosomal scale LD components for 5 representative cohorts (CEU, CHB, YRI, ASW, and 1KG). The upper parts of each figure represent (along the diagonal) and (off-diagonal), and the lower part as in Eq 8. For visualization purposes, the quantity of LD before scaling is transformed to a -log10 scale, with smaller values (red hues) representing larger LD, and a value of 0 representing that all SNPs are in LD. B) The relationship between the degree of population structure (approximated by ) and , and in the 26 1KG cohorts.

High-resolution illustration for LD grids for CEU, CHB, YRI, and ASW (m 250).

For each cohort, we partition chromosomes 6 and 11 into high-resolution LD grids (each LD grid contains 250 X 250 SNP pairs). The bottom half of each figure shows the LD grids for the entire chromosome. Further zooming into HLA on chromosome 6 and the centromere region on chromosome 11, and their detailed LD in the relevant regions are also provided in the upper half of each figure. For visualization purposes, LD is transformed to a -log10-scale, with smaller values (red hues) representing larger LD, and a value of 0 representing that all SNPs are in LD.

LD decay analysis for 26 1KG cohorts.

A) Conventional LD decay analysis in PLINK for 26 cohorts. To eliminate the influence of sample size, the inverse of sample size has been subtracted from the original LD values. The YRI cohort, represented by the orange dotted line, is chosen as the reference cohort in each plot. The top-down arrow shows the order of LDdecay values according to Table 5. B) Model-based LD decay analysis for the 26 1KG cohorts. We regressed each autosomal against its corresponding inversion of the SNP number for each cohort. Regression coefficient quantifies the averaged LD decay of the genome and intercept provides a direct estimate of possible existence of long-distance LD. The values in the first three plot indicate the correlation between and LD decay score in three different physical distance and the correlation between (left-side vertical axis) and LD decay score (right-side vertical axis) and the correlation between (left-side vertical axis) and (right-side vertical axis), respectively. The last plot assessed the impact of centromere region of chromosome 11 on the linear relationship between chromosomal LD and the inverse of the SNP number. The dark and light gray dashed lines represent the mean of the ℛwith and without the presence of centromere region of chromosome 11.

The correlation between the inversion of the SNP number and

A) The correlation between the inversion of the SNP number and in CEU, CHB, YRI, and ASW. B) Leave-one-chromosome-out strategy is adopted to evaluate the contribution of a certain chromosome on the correlation between the inverse of the SNP number and C) The correlation between the inversion of the SNP number and chromosomal LD in CEU, CHB, YRI, and ASW after removing the centromere region of chromosome 11. D) High-resolution illustration for LD grids for chromosome 8 in CEU, CHB, YRI, and ASW. For each cohort, we partition chromosome 8 into consecutive LD grids (each LD grid contains 250 X 250 SNP pairs). For visualization purposes, LD is transformed to a -log10-scale, with smaller values (red hues) representing larger LD, and a value of 0 representing that all SNPs are in LD.

The proposed algorithm has been realized in C++ and is available at: https://github.com/gc5k/gear2. As tested the software could handle sample size as large as 10,000 individuals.

Methods and Materials

The overall rationale for large-scale LD analysis

We assume LD for a pair of biallelic loci is measured by squared Pearson’s correlation, in which the LD of loci l1 and l2, p and q the reference and the alternative allele frequencies. If we consider the averaged LD for a genomic grid over SNP pairs, the conventional estimator is and, if we consider the averaged LD for mi and mj SNP pairs between two genomic segments, then . Now let us consider the 22 human autosomes (Figure 1A). We naturally partition the genome into 𝒞=22 blocks, and its genomic LD, denoted as ℓg can be expressed as So we can decompose ℓg into 𝒞 ℓi. and unique ℓi· j . Obviously, Eq 1 can be also expressed in the context for a single chromosome , in which the number of SNP segments, each of which has m SNPs. Geometrically it leads to βi diagonal grids and unique off-diagonal grids (Figure 1B).

LD-decay regression

As human genome can be boiled down to small LD blocks by genome-widely spread recombination hotspots (Hinch et al., 2019; Li et al., 2022), mechanically there is self-similarity for each chromosome that the relatively strong ℓi. for juxtaposed grids along the diagonal but weak ℓi· j for grids slightly off-diagonal. So, for a chromosomal ℓi., we can further express it as in which ℓu is the mean LD for a diagonal grid, ℓuv the mean LD for off-diagonal grids, and mi the number of SNPs on the ith chromosome. Consider a linear model below (Figure 1C for its illustration), in which ℓ represents a vector composed of 𝒞 ℓi., x represents a vector composed of 𝒞 xi and the inversion of the SNP number of the chromosome. The regression coefficient and intercept can be estimated as below And There are some technical details in order to find the interpretation for b0 and b1 .We itemize them briefly. For the mean and variance of x: For E(xℓ) For E(x) For E(ℓ) Then we integrate these items to have the expectation for b1: d

Similarly, we plug in E(b1) so as to derive b0 : After some algebra, if E(ℓu) ≫ E (ℓuv) – say if the former is far greater than the latter, the interpretation of b1 and b0 can be It should be noticed that E(b1) ≈ E(ℓu)m quantifies the averaged LD decay of the genome. Conventional LD decay is analysed via the well-known LD decay analysis, but Eq 4 provides a direct estimate of both LD decay and possible existence of extended LD. We will see the application of the model in Figure 5 that the strength of the long-distance LD is associated with population structure. of note, the underlying assumption of Eq 3 and Eq 4 is genome-wide spread of recombination hotspots, an established result that has been revealed and confirmed (Hinch et al., 2019).

Efficient estimation for ℓg, ℓi, and ℓi ·j

For the aforementioned analyses, the bottleneck obviously lies in the computational cost in estimating ℓi and ℓi·j· ℓi and ℓi·j are used to be estimated via the current benchmark algorithm as implemented in PLINK (Chang et al., 2015), and the computational time complex is proportional to O(nm2). We present a novel approach to estimate ℓi and ℓi·j· Given a genotypic matrix X, a n × m mmatrix, if we assume that there are mi and mj SNPs on chromosomes i and j, respectively, we can construct n × n genetic relatedness matrices as below in which is the standardized Xi and where xkl is the genotype for the Kth individual at the lth biallelic locus, F is the inbreeding coefficient having the value of 0 for random mating population and 1 for an inbred population, pl and ql are the frequencies of the reference and the alternative alleles (pl + ql =1)respectively. When GRM is given, we can obtain some statistical characters of Ki. We extract two vectors , which stacks the off-diagonal elements of Ki, and which takes the diagonal elements of Ki. The mathematical expectation of in which can be established according to Isserlis’s theorem in terms of the four-order moment (Isserlis, 1918), in which is the expected relatedness score and r indicates the rth degree relatives. r=0 for the same individual, and r=1 for the first degree relatives. Similarly, we can derive for Eq 6 establishes the connection between GRM and the aggregated LD estimation that .According to Delta method as exampled in Appendix I of Lynch and Walsh’s book (Lynch and Walsh, 1998), the means and the sampling variances for ℓi and ℓi·j are, in which and , respectively. Of note, the properties of ℓg can be derived similarly if we replace ℓi with ℓg in Eq 7. We can develop , a scaled version of ℓi·j· as below in which , a modification that removed the LD with itself. According to Delta method, the sampling variance of is Of note, when there is no LD between a pair of loci, ℓ yields zero and its counterpart PLINK estimate yields ,a difference that can be reconciled in practice (see Figure 2).

Raise of LD due to population structure

In this study, the connection between LD and population structure is bridged via two pathways below, in terms of a pair of loci and of the aggregated LD for all pair of loci. For a pair of loci, their LD is often simplified as , but will be inflated if there are subgroups (Nei and Li, 1973). In addition, it is well established the connection between population structure and eigenvalues, and in particular the largest eigenvalue is associated with divergence of subgroups (Patterson et al., 2006). In this study, the existence of subgroups of cohort is surrogated by the largest eigenvalue .

Data description and quality control

The 1KG (Auton et al., 2015), which is launched to produce a deep catalogue of human genomic variation by whole genome sequencing (WGS) or whole exome sequencing (WES), and 2,503 strategically selected individuals of global diversity are included (containing 26 cohorts). We used the following criteria for SNP inclusion for each of the 26 1KG cohorts: i) autosomal SNPs only; ii) SNPs with missing genotype rates higher than 0.2 were removed, and missing genotypes were imputed; iii) Only SNPs with minor allele frequencies higher than 0·05 were retained. Then 2,997,635 consensus SNPs that were present in each of the 26 cohorts were retained. According to their origins, the 26 cohorts are grouped as African (AFR: MSL, GWD, YRI, ESN, ACB, LWK, and ASW), European (EUR: TSI, IBS, CEU, GBR, and FIN), East Asian (EA: CHS, CDX, KHV, CHB, and JPT), South Asian (SA: BEB, ITU, STU, PJL, and GIH), and American (AMR: MXL, PUR, CLM, and PEL), respectively.

In addition, to test the capacity of the developed software (X-LD), we also included CONVERGE cohort (n=10,640), which was used to investigate Major Depressive Disorder (MDD) in the Han Chinese population (Cai et al., 2015). We performed the same criteria for SNP inclusion as that of the 1KG cohorts, and m=5,215,820 SNPs were remained for analyses.

X-LD software implementation

The proposed algorithm has been realized in our X-LD software, which is written in C++ and reads in binary genotype data as often used in PLINK. As multi-thread programming is adopted, the efficiency of X-LD can be improved upon the availability of computational resources. We have tested X-LD in various independent datasets for its reliability and robustness. Certain data management options, such as flexible inclusion or exclusion of chromosomes, have been built into the commands of X-LD. In X-LD, missing genotypes are naively imputed according to Hardy-Weinberg proportions; however, when missing rate is high, we suggest the genotype matrix should be imputed by other advanced imputation tools.

The most time-consuming part of X-LD was the construction of GRM , and the established computational time complex was 𝒪 (n2m). However, if is decomposed into , in which , has dimension of n × B, using Mailman algorithm the computational time complex for building K can be reduced to (Liberty and Zucker, 2009). This idea of embedding Mailman algorithm into certain high throughput genomic studies has been successful, and our X-LD software is also leveraged by absorbing its recent practice in genetic application (Wu and Sankararaman, 2018).

Results

Statistical properties of the proposed method

Table 1 introduced symbols frequently cited in this study. As schematically illustrated in Figure 1, ℓg could be decomposed into 𝒞 ℓi and unique ℓi,j components. We compared the estimated ℓ;i and ℓi,j in X-LD with those being estimated in PLINK (known as “--r2”, and the estimated squared Pearson’s correlation LD is denoted as r2). Considering the 212 substantial computational cost of PLINK, only 100,000 randomly selected autosome SNPs were used for each 1KG cohort, and 22 and 231 were estimated. After regressing 22 against those of PLINK, we found that the regression slope was close to unity and bore an anticipated intercept a quantity of approximately (Figure 2A and Figure 2B).

In other words, PLINK gave even for SNPs of no LD. However, when regressing 231 estimates against those of PLINK, it was found that largely because of tiny quantity of it was slightly smaller than 1 but statistically insignificant from 1 in these 26 1KG cohorts (mean of 0.86 and s.d. of 0.10, and its 95 % confidence interval was (0.664, 1.056)); when the entire 1KG samples were used, its much larger LD due to subgroups, nearly no estimation bias was found (Figure 2A and Figure 2B). In contrast, because of their much larger values, components were always consistent with their corresponding estimates from PLINK (mean of 1.03 and s.d. of 0.012, 95% confidence interval was (1.006, 1.053), bearing an ignorable bias). Furthermore, we also combined the African cohorts together (MSL, GWD, YRI, ESN, LWK, totaling 599 individuals), the East Asian cohorts together (CHS, CDX, KHV, CHB, and JPT, totaling 504 individuals), and the European cohorts together (TSI, IBS, CEU, GBR, and FIN, totaling 503 individuals), the resemblance pattern between X-LD and PLINK was similar as observed in each cohort alone (Figure S1). The empirical data in 1KG verified that the proposed method was sufficiently accurate.

To fairly evaluate the computational efficiency of the proposed method, the benchmark comparison was conducted on the first chromosome of the entire 1KG dataset (n =2.503) and m =(225,967), and 10 CPUs were used for multi-thread computing. Compared with PLINK, the calculation efficiency of X-LD was nearly 30∼40 times faster for the tested chromosome, and its computational time of X-LD was proportional to (Figure S2). So, X-LD provided a feasible and reliable estimation of large-scale complex LD patterns. More detailed computational time of the tested tasks would be reported in their corresponding sections below; since each 1KG cohort had sample size around 100, otherwise specified the computational time was reported for CHB (n = 103) as a reference (Table 2). In order to test the capability of the software, the largest dataset tested was CONVERGE (n =10,640, and (n =5,215,820), and it took 77,508.00 seconds, about 22 hours, to estimate 22 autosomal and 231 (Figure 1A);

When zooming into chromosome 2 of CONVERGE, on which 420,949 SNP had been evenly split into 1,000 blocks and yielded 1,000 grids, and 499,500 LD grids, it took 45,125.00 seconds, about 12.6 hours, to finished the task (Figure 1B).

Ubiquitously extended LD and population structure/admixture

We partitioned the 2,997,635 SNPs into 22 autosomes (Figure 3A and Figure S3), and the general LD patterns were as illustrated for CEU, CHB, YRI, ASW, and 1KG. As expected, for each cohort (Figure 3B). As observed in these 1KG cohorts, all these three LD measures were associated with population structure, which was surrogated by , and their squared correlation R2 were greater than 0.8. ACB, ASW, PEL, and MXL, which all showed certain admixture, tended to have much greater , and (Table 3 and Figure 3B). In contrast, East Asian (EA) and European (EUR) orientated cohorts, which showed little within cohort genetic differentiation – as their largest eigenvalues were slightly greater than 1, had their aggregated LD relatively low and resembled each other (Table 3). Furthermore, for several European (TSI, IBS, and FIN) and East Asian (JPT) cohorts, the ratio between and components could be smaller than 0.1, and the smallest ratio was found to be about 0.091 in FIN. The largest ratio was found in 1KG that and , and the ratio was 0.877 because of the inflated LD due to population structure. A more concise statistic to describe the ratio between ℓi,jand ℓ;i was , Eq 8, and the corresponding values for 231 scaled for FIN was (s.d. of 0.027) and for 1KG was (s.d. of 0.028).

In terms of computational time, for 103 CHB samples, it took about 101.34 seconds to estimate 22 autosomal and 231 ; for all 1KG 2,503 samples, X-LD took about 3,008.29 seconds (Table 1). Conventional methods took too long to complete the analyses in this section, so no comparable computational time was provided. For detailed 22 and 231 estimates for each 1KG cohort, please refer to Extended Data 1 (Excel Sheet 1-27).

Detecting exceedingly high LD grids shaped by variable recombination rates

We further explored each autosome with high-resolution grid LD visualization. We set m =250, so each grid had the ℓ;uv for 250 × 250 SNP pairs. The computational time complex was , in which , and with our proposed method in CHB it costed 66.86 seconds for chromosome 2, which had 241,241 SNPs and was totaled 466,095 unique grids, and 3.22 seconds only for chromosome 22, which had 40,378 SNPs and was totaled 13,203 unique grids (Table 1). In contrast, under conventional methods those LD grids were not very likely to be exhaustively surveyed because of its computational cost was : for CHB chromosome 2, it would have taken about 40 hours as estimated. As the result was very similar for m =500 (Figure S4), we only reported the results under m =250 below.

As expected, chromosome 6 (206,165 SNPs, totaling 340,725 unique grids) had its HLA cluster showing much higher LD than the rest of chromosome 6. In addition, we found very dramatic variation of the HLA cluster LD HLA (28,477,797-33,448,354 bp, totaling 3,160 unique grids) across ethnicities. For CEU, CHB, YRI, and ASW, their , and 0.0019, respectively, but their corresponding HLA cluster grids had and 0.022, respectively (Figure 4).

Consequently, the largest ratio for was of 42.00 in CEU, 39.06 in YRI, and 32.22 in CHB, but was reduced to 11.58 in ASW. Before the release of CHM13 (Hoyt et al., 2022), chromosome 11 had the most completely sequenced centromere region, which had much rarer recombination events, all four cohorts showed an strong LD around the centromere (46,061,947-59,413,484 bp, totaling 1,035 unique grids) regardless of their ethnicities (Figure 4). , and 0.0022, respectively, and , and 0.094, respectively; the ratio for , and 94,05, for CEU, CHB, and YRI, respectively; the lowest ratio was found in ASW of 42.73. In addition, removing the HLA region of chromosome 6 or the centromere region of chromosome 11 would significantly reduce or in comparison with the randomly removal of other regions (Figure S5).

Model-based LD decay regression revealed LD composition

The real LD block size was not exact of m =250 or m =500, but an unknown parameter that should be inferred in computational intensive “LD decay” analysis (Zhang et al., 2019; Chang et al., 2015). We conducted the conventional LD decay for the 26 1KG cohorts (Figure 5A), and the time cost was 1,491.94 seconds for CHB. For each cohort, we took the area under the LD decay curve in the LD decay plot, and it quantified approximately the LD decay score for each cohort. The smallest score was 0.0421 for MSL and the largest was 0.0598 for PEL (Table 5). However, this estimation was not taken into account the real extent of LD, so it was not precise enough to reflect the LD decay score. For example, for admixture population, such as American cohorts, the extent of LD would be longer.

In contrast, we proposed a model-based method, as given in Eq 3, which could estimate LD decay score (regression coefficient b1) and long-distance LD score (intercept b0) jointly. Given the estimated 22 (Extended data 1; Table 4 for four representative cohorts and Supplementary R code), we regressed each autosomal against its correspondingly inversion of SNP number, and all yielded positive slopes (Pearson’s correlation ℛ > 0.80, Table 5; Figure 5B), an observation that was consistent with genome-wide spread of recombination hotspots. This linear relationship could consequently be considered the norm for a relative homogenous population as observed in most 1KG cohorts (Figure S6), while for the all 2,503 1KG samples ℛ =0.55 only (Table 5), indicating that the population structure and possible differentiated recombination hotspots across ethnicities disturbed the assumption underlying Eq 3 and smeared the linearity. We extracted and for the 26 1KG cohorts for further analysis. The rates of LD decay score, as indicated by , within the African cohorts (AFR) were significantly faster than other continents, consistent with previous observation that African population had relative shorter LD (Gabriel et al., 2002); while subgroups within the American continent (AMR) tended to have extended LD range due to their admixed genetic composition (Table 5 and Figure 5B). Notably, the correlation between and the approximated LD decay score was ℛ > 0.88. The estimated were highly correlated with .

A common feature was universally relative high LD of chromosome 6 and 11 in the 26 1KG cohorts (Figure S6). We quantified the impact of chromosome 6 and 11 by leave-one-chromosome-out test in CEU, CHB, YRI, and ASW for details (Figure 6A and 6B), and found that dropping chromosome 6 off could lift ℛ on average by 0.017, and chromosome 11 by 0.046. One possible explanation was that the centromere regions of chromosomes 6 and 11 have been assembled more completely than other chromosomes before the completion of CHM13 (Hoyt et al., 2022), whereas meiotic recombination tended to be reduced around the centromeres (Hinch et al., 2019). We estimated ℓ;i after having knocked out the centromere region (46,061,947-59,413,484 bp, chr 11) in CEU, CHB, YRI, and ASW, and chromosome 11 then did not deviate much from their respective fitted lines (Figure 6C). A notable exceptional pattern was found in ASW, the chromosome 8 of which had even more deviation than chromosome 11 (ℛ was 0.83 and 0.87 with and without chromosome 8 in leave-one-chromosome out test) (Figure 6B). The deviation of chromosome 8 of ASW was consistent even more SNPs were added (Figure S7). We also provided high-resolution LD grids illustration for chromosome 8 (163,436 SNPs, totaling 214,185 grids) of the four representative cohorts for more detailed virtualization (Figure 6D). ASW had , but 0.00075, 0.00069 and 0.00043 for CEU, CHB, and YRI, respectively.

Discussion

In this study, we present a computationally efficient method to estimate mean LD of genomic grids of many SNP pairs. Our LD analysis framework is based on GRM, which has been embedded in variance component analysis for complex traits and genomic selection (Goddard, 2009; Visscher et al., 2014; Chen, 2014). The key connection from GRM to LD is bridged via the transformation between n × n matrix and m × m matrix, in particular here via Isserlis’s theorem under the fourth-order moment (Isserlis, 1918). With this connection, the computational cost for estimating the mean LD of m × m SNP pairs is reduced from 𝒪(nm2) to 𝒪(n2m), and the statistical properties of the proposed method are derived in theory and validated in 1KG datasets. In addition, as the genotype matrix X is of limited entries {0, 1, 2}, assuming missing genotypes are imputed first, using Mailman algorithm the computational cost of GRM can be further reduced to (Liberty and Zucker, 2009). The largest data tested so far for the proposed method has the sample size of 10,640 and of more than 5 million SNPs, it can complete genomic LD analysis in 77,508.00 seconds (Table 1). The weakness of the proposed method is obvious that the algorithm remains slow when the sample size is large or the grid resolution is increased. With the availability of such as UK Biobank data (Bycroft et al., 2018), the proposed method may not be adequate, and much advanced methods, such as randomizated implementation for the proposed methods, are needed.

We also applied the proposed method into 1KG and revealed certain characteristics of the human genomes. Firstly, we found the ubiquitously existence of extended LD, which was likely emerged because of population structure, even very slightly, and admixture history. We quantified the and in 1KG, and as indicated by we found the inter-chromosomal LD was nearly an order lower than intra-chromosomal LD; for admixed cohorts, the ratio was much higher, even very close to each other such as in all 1KG samples. Secondly, variable recombination rates shaped peak of local LD. For example, the HLA region showed high LD in European and East Asian cohorts, but relatively low LD in such as YRI, consistent with their much longer population history. Thirdly, it existed general linear correlation between ℓ;i and the inversion of the SNP number, a long-anticipated result that is as predicted with genome-wide spread of recombination hotspots (Hinch et al., 2019). One outlier of this linear norm was chromosome 11, which had so far most completely genotyped centromere and consequently had more elevated LD compared with other autosomes. We anticipate that with the release of CHM13 the linear correlation should be much closer to unity (Hoyt et al., 2022). Of note, under the variance component analysis for complex traits, it is often a positive correlation between the length of a chromosome (as surrogated by the number of SNPs) and the proportion of heritability explained (Chen et al., 2014).

In contrast, throughout the study recurrent outstanding observations were found in ASW. For example, in ASW the ratio of was substantially dropped down compared with that of CEU, CHB, or YRI as illustrated in Figure 4. Furthermore, chromosome 8 in ASW fluctuated upwards most from the linear correlation (Figure 6), and even after various analyses, such as expanding SNP numbers. One possible explanation may lay under the complex demographic history of ASW, which can be investigated and tested in additional African American samples or possible existence for epistatic fitness (Ni et al., 2020).

Supporting information